API design for machine learning software: experiences from the scikit-learn project

### 2013-10-21 paper.tex
some more copy-editing/trimming

\begin{description}
  \item[Consistency.]
       All objects (basic or composite) share a consistent interface composed of
       a limited set of methods. This interface is documented in a consistent
       manner for all objects.
  \item[Inspection.]
       Constructor parameters and parameter values determined by learning
       algorithms are stored and exposed as public attributes.
  \item[Non-proliferation of classes.]
       Learning algorithms are the only objects to be represented using custom classes.
       Datasets are represented as NumPy arrays or SciPy sparse matrices.
       Hyper-parameter names and values are represented as standard
       Python strings or numbers whenever possible.
       This keeps \sklearn easy to use and easy to combine with other libraries.
  \item[Composition.]
       Learning algorithms can be combined into processing pipelines.
       Ensemble methods and other meta-algorithms
       are implemented as adapter classes.
       User-defined algorithms that follow the right conventions
       can be used as building blocks in the resulting composite algorithms.
  \item[Sensible defaults.]
       Whenever an operation requires a user-defined parameter,
       an appropriate default value is defined by the library.
       The default value should cause the operation to be performed
       in a sensible way (giving a baseline solution for the task at hand).
\end{description}

@@

To illustrate the initialize-fit sequence,
let us consider a supervised learning task using logistic regression.
Given the API defined above, solving this problem is as simple as the following
example.
\begin{footnotesize}
\begin{pythoncode}
from sklearn.linear_model import LogisticRegression
clf = LogisticRegression(penalty="l1")
clf.fit(X_train, y_train)
\end{pythoncode}
\end{footnotesize}
In this snippet, a \texttt{LogisticRegression} estimator is first initialized by
setting the \texttt{penalty} hyper-parameter to \texttt{"l1"} for
$\ell_1$ regularization. Other hyper-parameters (such as \texttt{C},
the strength of the regularization) are not explicitly given and
thus set to the default values. Upon calling \texttt{fit}, a model is
learned from the training arrays \texttt{X\_train} and \texttt{y\_train},
and stored on the object for later use.
%% (In this example, \texttt{fit} is called solely for its side effects.)
Since all estimators share the same interface, using a different learning algorithm is
as simple as replacing the constructor (the class name).
To build a random forest on
the same data, one would simply replace
\texttt{LogisticRegression(penalty="l1")} in the snippet above by
\texttt{RandomForestClassifier()}.

In \sklearn, classical learning algorithms are not the only objects to be
implemented as estimators. For example, preprocessing routines (e.g., scaling of
features) or feature extraction techniques (e.g., vectorization of text
documents) also implement the \textit{estimator} interface. Even stateless
processing steps, that do not require the \texttt{fit} method to
perform useful work, implement the estimator interface. As we will illustrate
in the next sections, this design pattern is indeed of prime importance for
consistency, composition and model selection reasons.

@@

Since it is common to modify or filter data before feeding it to a learning
algorithm, some estimators in the library implement a \textit{transformer}
interface which defines a \texttt{transform} method. It takes as input some new
data \texttt{X\_test} and yields as output a transformed version of
\texttt{X\_test}. Preprocessing, feature selection, feature extraction and dimensionality reduction
algorithms are all provided as transformers.  In our example,
to standardize the input \texttt{X\_train} to zero mean and unit variance
before fitting the logistic regression estimator, one would write:
\begin{small}
\begin{pythoncode}
from sklearn.preprocessing import StandardScaler
scaler = StandardScaler()
scaler.fit(X_train)
X_train = scaler.transform(X_train)
\end{pythoncode}
\end{small}
Of course, in practice, it is important to apply the same preprocessing to the
test data \texttt{X\_test}. Since a \texttt{StandardScaler} estimator stores the
mean and standard deviation that it computed for the training set, transforming
an unseen test set \texttt{X\_test} maps it into the appropriate region of
feature space:
\begin{pythoncode}
X_test = scaler.transform(X_test)
\end{pythoncode}
Transformers also include a variety of learning algorithms, such as
dimensionality reduction (PCA, manifold learning), kernel approximation,
and other mappings from one feature space to another.

@@

As introduced in Section~\ref{sec:estimators}, hyper-parameters set in the
constructor of an estimator
determine the behavior of the learning algorithm
and hence the performance of the resulting model on unseen data.
For example, a decision
tree with too small a value for the maximal tree depth
parameter will tend to underfit, while too large a value will make it overfit.
The problem of \textit{model selection} is therefore to find, within
some hyper-parameter space, the best combination of hyper-parameters, with
respect to some user-specified criterion.

In \sklearn, model selection is supported by the meta-estimators
\texttt{GridSearchCV} and \texttt{RandomizedSearchCV}.  These take as input an
estimator (basic or composite), whose hyper-parameters must be optimized, and a
set of hyperparameter settings to search through.
This set is represented as a mapping of parameter names
to a set of discrete choices in the case of grid search,
which exhaustively enumerates the ``grid'' (cartesian product)
of complete parameter combinations.
Randomized search is a more efficient algorithm
that avoids the combinatorial explosion in grid search
by sampling a fixed number of times from its parameter distributions
\citep{bergstra2012}.

@@

The code in Figure~\ref{svm-gridsearch} shows how to find
hyper-parameter settings for a support vector classifier (SVC)
that maximizes $F_1$ score
through 10-fold cross-validation on the training set.
Two distinct hyper-parameter grids are
considered for the linear and radial basis function (RBF) kernels;
an SVC with a linear kernel accepts a $\gamma$ parameter, but ignores it,
so using a single parameter grid would waste computing time
trying out effectively equivalent settings.

@@

To facilitate the installation of \sklearn,
the set of external dependencies is kept to a bare minimum:
only Python, NumPy and SciPy are required for a functioning installation.
Binary distributions of these are available for the major platforms.
Visualization functionality uses Matplotlib \citep{hunter2007matplotlib}
and Graphviz \citep{gansner2000},
but neither is required to perform machine learning or prediction.
When feasible, external libraries are integrated into the codebase.
In particular, \sklearn includes modified versions of \textsf{LIBSVM} and \textsf{LIBLINEAR}
\citep{chang2011libsvm,fan2008}, both written in C{}\verb!++!
and wrapped using Cython.

@@

Other machine learning packages,
such as SofiaML\footnote{\url{https://code.google.com/p/sofia-ml}}
and Vowpal~Wabbit,\footnote{\url{http://hunch.net/~vw}}
are often designed as command-line tools
and may not offer any kind of API\@.
While this has the advantage that users are not tied
to a particular programming language,
they will find that they still need programming to process input/output,
and will do so in a variety of languages.
By contrast, \sklearn allows users to implement that entire workflow
in a single program, written in a single language.
This also makes it easier for researchers and developers
to exchange and collaborate on software, as dependencies and setup are kept to a
minimum.

Similar benefits hold in the case of specialized languages
for numeric and statistical programming
such as \textsc{matlab} and R \citep{r}.
In comparison to these, though, Python has the distinct advantage
that it is a \textit{general purpose} language,
while NumPy and SciPy extend it with functionality
similar to that offered by \textsc{matlab} and R.
Python has strong language and standard library support for such tasks as
string/text processing, interprocess communication, networking
and many of the other auxiliary tasks that machine learning programs
(whether academic or commercial) routinely need to perform.
While support for many of these tasks is improving in languages such as
\textsc{matlab} and R, they still lag behind Python in their general purpose
applicability.
In many applications of machine learning, these tasks, such as data access,
data preprocessing and reporting, can take more work than applying
the actual learning algorithm.

@@

Finally, the Gensim topic modeling toolkit \citep{rehurek2010gensim}
exemplifies an API geared toward ``big data'' processing.
Gensim's method of dealing with large datasets is to use algorithms
that have small space complexity and can be updated online.
The API is designed around the Python concept of an ``iterable''
(supported in the language by a restricted form of co-routines called generators).
Some \sklearn estimators, among which the implementations of naïve Bayes
and minibatch $k$-means,
have a \texttt{partial\_fit} method to learn from a stream of samples
that is divided into mini-batches (chunks).
This way of processing, as argued in Section~\ref{sec:arrays},
allows for more efficient implementation.
We make no attempt to hide this batch-oriented processing from the user,
allowing control over the amount of memory dedicated to \sklearn algorithms.

@@

At a lower-level, parallel processing is a potential point of improvement.
Some estimators in \sklearn are already able to leverage multicore processors,
but only in a coarse-grained fashion.
Parallel processing is difficult to accomplish in the Python environment;
\sklearn targets the main implementation, CPython,
which cannot execute Python code on multiple CPUs simultaneously.
It follows that any parallel task decomposition must either be done
inside Cython modules,
or at a level high enough to warrant the overhead
of creating multiple OS-level processes,
and the ensuing inter-process communication.
Parallel grid search is an example of the latter approach
which has already been implemented.
Experiments are underway to implement more fine-grained multicore support.

>>>>

\begin{description}
  \item[Consistency.]
       All objects (basic or composite) share a consistent interface composed of
       a limited set of methods. This interface is documented in a consistent
       manner for all objects.
  \item[Inspectability.]
       Parameters, whether set by the user or learned from data,
       are stored and exposed as public attributes.
  \item[Non-proliferation of classes.]
       Learning algorithms are the only objects to be represented using custom classes.
       Datasets are represented as NumPy arrays or SciPy sparse matrices.
       User-set parameter names and values are represented as standard
       Python strings or numbers, not custom objects.
       This keeps \sklearn easy to use and easy to combine with other libraries.
  \item[Composability.]
       Learning algorithms can be combined into processing pipelines.
       Ensemble methods and other meta-algorithms
       are implemented as adapter classes.
       User-defined algorithms that follow the right conventions
       can be used as building blocks in the resulting composite algorithms.
  \item[Sensible defaults.]
       Whenever an operation requires a user-defined parameter,
       a sensible default value is defined by the library
       to give a baseline solution for the task at hand.
\end{description}

@@

To illustrate the initialize-fit sequence,
let us consider a supervised learning task using logistic regression.
Given the API defined above, solving this problem is as simple as the following
example.
\begin{footnotesize}
\begin{pythoncode}
from sklearn.linear_model import \
    LogisticRegression
clf = LogisticRegression(penalty="l1")
clf.fit(X_train, y_train)
\end{pythoncode}
\end{footnotesize}
In this snippet, a \texttt{LogisticRegression} estimator is first initialized by
setting the \texttt{penalty} hyper-parameter to \texttt{"l1"} for
$\ell_1$ regularization. Other hyper-parameters (such as \texttt{C},
the strength of the regularization) are not explicitly given and
thus set to the default values. Upon calling \texttt{fit}, a model is
learned from the training arrays \texttt{X\_train} and \texttt{y\_train},
and stored on the object for later use.
%% (In this example, \texttt{fit} is called solely for its side effects.)
Since all estimators share the same interface, using a different learning algorithm is
as simple as replacing the constructor (the class name).
To build a random forest on
the same data, one would simply replace
\texttt{LogisticRegression(penalty="l1")} in the snippet above by
\texttt{RandomForestClassifier()}.

In \sklearn, classical learning algorithms are not the only objects to be
implemented as estimators. For example, preprocessing routines (e.g., scaling of
features) or feature extraction techniques (e.g., vectorization of text
documents) also implement the \textit{estimator} interface. Even stateless
processing steps, that do not require the \texttt{fit} method to
perform useful work, implement the estimator interface. As we will illustrate
in the next sections, this design pattern is of prime importance for
consistency, composition and model selection reasons.

@@

Since it is common to modify or filter data before feeding it to a learning
algorithm, some estimators in the library implement a \textit{transformer}
interface which defines a \texttt{transform} method. It takes as input some new
data \texttt{X\_test} and yields as output a transformed version of
\texttt{X\_test}. Preprocessing, feature selection, feature extraction and dimensionality reduction
algorithms are all provided as transformers.  In our example,
to standardize the input \texttt{X\_train} to zero mean and unit variance
before fitting the logistic regression estimator, one would write:
\begin{small}
\begin{pythoncode}
from sklearn.preprocessing import StandardScaler
scaler = StandardScaler()
scaler.fit(X_train)
X_train = scaler.transform(X_train)
\end{pythoncode}
\end{small}
To obtain good results, it is necessary to apply the exact same preprocessing to the
test data \texttt{X\_test}.
For this reason, a \texttt{StandardScaler} stores the mean and standard deviation
that it computed for the training set
and will use these when asked to \texttt{transform}
an unseen set of samples \texttt{X\_test}:

Transformers also include a variety of learning algorithms, such as
dimensionality reduction, manifold learning, kernel approximation,
and other mappings from one feature space to another.

@@

Hyper-parameters values set in the constructor of an estimator
(Section~\ref{sec:estimators})
determine the behavior of the learning algorithm
and the performance of the resulting model on unseen data.
The problem of \textit{model selection} is to find
the optimal combination of hyper-parameters, with
respect to some user-specified criterion. For example, a decision
tree with too small a maximal tree depth parameter will fit its training set badly,
while too large a value will make it overfit.

In \sklearn, model selection is supported by the meta-estimators
\texttt{GridSearchCV} and \texttt{RandomizedSearchCV}.  These take as input an
estimator, basic or composite,
and a set of hyperparameter settings to search through.
This set is represented as a mapping of parameter names
to a set of discrete choices in the case of grid search,
which exhaustively enumerates the ``grid'' (cartesian product)
of complete parameter combinations.
Randomized search is a more efficient algorithm
that avoids the combinatorial explosion in grid search
by sampling a fixed number of times from its parameter distributions
\citep{bergstra2012}.

@@

Figure~\ref{svm-gridsearch} shows code that finds
settings for an SVM classifier (SVC)
that maximize $F_1$ score
through 10-fold cross-validation.
Two distinct hyper-parameter grids are
considered for the linear and radial basis function (RBF) kernels:
an SVM with a linear kernel accepts a $\gamma$ parameter, but ignores it,
so using a single grid would waste computing time
trying out effectively equivalent settings.

@@

To facilitate the installation of \sklearn,
the set of external dependencies is kept to a bare minimum:
only Python, NumPy and SciPy are required for a functioning installation.
Binary distributions of these are available for the major platforms.
Visualization functionality uses Matplotlib \citep{hunter2007matplotlib}
and Graphviz \citep{gansner2000},
but neither is required to perform machine learning or prediction.
When feasible, external libraries are integrated into the codebase.
In particular, \sklearn includes modified versions of the C{}\verb!++! libraries
\textsf{LIBSVM} and \textsf{LIBLINEAR} \citep{chang2011libsvm,fan2008}.

@@

Other machine learning packages,
such as SofiaML\footnote{\url{https://code.google.com/p/sofia-ml}}
and Vowpal~Wabbit,\footnote{\url{http://hunch.net/~vw}}
are often designed as command-line tools without any kind of API\@.
While this has the advantage that users are not tied
to a particular programming language,
they will find that they still need programming to process input/output,
and will do so in a variety of languages.
By contrast, \sklearn allows users to implement that entire workflow
in a single language.
This makes it easier for researchers and developers
to exchange and collaborate on software, as dependencies and setup are kept to a
minimum.

Similar benefits hold in the case of specialized languages
for numeric and statistical programming
such as \textsc{matlab} and R \citep{r}.
In comparison to these, though, Python has the distinct advantage
that it is a \textit{general-purpose} language,
while NumPy and SciPy extend it with functionality
similar to that offered by \textsc{matlab} and R.
Python has strong language and standard library support for such tasks as
string/text processing, interprocess communication, networking
and many of the other auxiliary tasks that machine learning programs
(whether academic or commercial) routinely need to perform.
While support for many of these tasks is improving in languages such as
\textsc{matlab} and R, they still lag behind Python in this regard.
In many applications of machine learning, data access,
preprocessing, and reporting can take more work than applying
the actual learning algorithm.

@@

Finally, the Gensim topic modeling toolkit \citep{rehurek2010gensim}
exemplifies an API geared toward ``big data'' processing.
Gensim's method of dealing with large datasets is to use algorithms
that have small space complexity and can be updated online.
The API is designed around the Python concept of an ``iterable''
(supported in the language by a restricted form of co-routines called generators).
Some \sklearn estimators, including the implementations of naïve Bayes
and minibatch $k$-means,
have a \texttt{partial\_fit} method to learn from a stream of samples
that is divided into mini-batches (chunks).
This way of processing, as argued in Section~\ref{sec:arrays},
allows for efficient implementation.
We make no attempt to hide this batch-oriented processing from the user,
allowing control over the amount of memory dedicated to \sklearn algorithms.

@@

At a lower-level, parallel processing is a potential point of improvement.
Some estimators in \sklearn are already able to leverage multicore processors,
but only in a coarse-grained fashion.
Parallel processing is difficult to accomplish in the Python environment;
\sklearn targets the main implementation, CPython,
which cannot execute Python code on multiple CPUs simultaneously.
It follows that parallel task decomposition must either be done
inside Cython modules,
or at a level high enough to warrant the overhead
of inter-process communication.
Parallel grid search is an example of the latter approach
which has already been implemented.
Experiments are underway to implement more fine-grained multicore support.
